---
title: "Importing data into R and understanding different data types - notebook"
author: "Paul Bradshaw"
date: "10/08/2020"
output: html_document
---

# Importing data into R and understanding different data types - notebook

This notebook details: 

* How to read data into RStudio from a CSV
* Functions
* Variables
* 'Object' types, such as text strings, numbers, tables and lists
* Basic analysis and calculations

## Reading data into RStudio from a CSV

We've downloaded a CSV file from the gender pay gap service's [Download gender pay gap data](https://gender-pay-gap.service.gov.uk/viewing/download) page. There is a file for every year - we've picked the [2019-20 data](https://gender-pay-gap.service.gov.uk/viewing/download-data/2019) because at the time of writing that's the last full year of data.

Once downloaded, it's important to **put the file in the same location as your R project**.

To import a CSV into R, you need to use a **function** - in this case, we are going to use one called `read.csv()`.

Functions are like recipes: a series of instructions for achieving a certain result. But instead of having to type all those instructions, you just use the name of the function (in the same way as you might say "use the cookie recipe" instead of giving someone a list of instructions each time). 

The `read.csv()` function, for example, is, not surprisingly, a recipe for reading CSV files.

Like recipes, functions normally have **ingredients**. For that reason, a function is always followed by a pair of parentheses, containing any ingredients that it needs (again, in the same way as you might say "use the cookie recipe, but add chocolate chips").

The term for those ingredients in R is ["parameters" or "arguments"](https://stackoverflow.com/questions/15463851/confused-about-r-terminology-attributes-parameters-and-arguments): when we talk about a function in general, for example, we might say it "takes two arguments" (two ingredients). 

When a function is used, then, it might look something like this: 

`myfunction(the_first_argument)`

If it has more than one ingredient then each argument is separated by a comma, like so:

`myfunction(the_first_argument, the_second_argument)`

The code below, then, uses the name of the CSV file as the *argument* for the function `read.csv()`.

```{r}
read.csv("UK Gender Pay Gap Data - 2019 to 2020.csv")
```

When that code is run, the CSV is used as the one ingredient for the function `read.csv()`, and the results - thousands of rows of data - are shown. 

Notice that the name of the file is in **quotation marks**: it has to be in quotation marks or the code won't work. We will come onto why that's the case later, but for now all you need to know is if you're specifying the name or location of a file on your machine then you need to put it inside quotation marks.

But when we read that data in using `read.csv()`, we haven't actually *stored* it to be able to work with.

To do that we need an extra step:

```{r}
genderpay1920 <- read.csv("UK Gender Pay Gap Data - 2019 to 2020.csv")
```

