---
title: "Importing data into R and understanding different data types - notebook"
author: "Paul Bradshaw"
date: "10/08/2020"
output: html_document
---

# Importing data into R and understanding different data types - notebook

This notebook details: 

* How to read data into RStudio from a CSV
* Functions
* Variables
* 'Object' types, such as text strings, numbers, tables and lists
* Basic analysis and calculations

## Reading data into RStudio from a CSV

We've downloaded a CSV file from the gender pay gap service's [Download gender pay gap data](https://gender-pay-gap.service.gov.uk/viewing/download) page. There is a file for every year - we've picked the [2019-20 data](https://gender-pay-gap.service.gov.uk/viewing/download-data/2019) because at the time of writing that's the last full year of data.

Once downloaded, it's important to **put the file in the same location as your R project**.

To import a CSV into R, you need to use a **function** - in this case, we are going to use one called `read.csv()`.

Functions are like recipes: a series of instructions for achieving a certain result. But instead of having to type all those instructions, you just use the name of the function (in the same way as you might say "use the cookie recipe" instead of giving someone a list of instructions each time). 

The `read.csv()` function, for example, is, not surprisingly, a recipe for reading CSV files.

Like recipes, functions normally have **ingredients**. For that reason, a function is always followed by a pair of parentheses, containing any ingredients that it needs (again, in the same way as you might say "use the cookie recipe, but add chocolate chips").

The term for those ingredients in R is ["parameters" or "arguments"](https://stackoverflow.com/questions/15463851/confused-about-r-terminology-attributes-parameters-and-arguments): when we talk about a function in general, for example, we might say it "takes two arguments" (two ingredients). 

When a function is used, then, it might look something like this: 

`myfunction(the_first_argument)`

If it has more than one ingredient then each argument is separated by a comma, like so:

`myfunction(the_first_argument, the_second_argument)`

The code below, then, uses the name of the CSV file as the *argument* for the function `read.csv()`.

```{r}
read.csv("UK Gender Pay Gap Data - 2019 to 2020.csv")
```

When that code is run, the CSV is used as the one ingredient for the function `read.csv()`, and the results - thousands of rows of data - are shown. 

Notice that the name of the file is in **quotation marks**: it has to be in quotation marks or the code won't work. We will come onto why that's the case later, but for now all you need to know is if you're specifying the name or location of a file on your machine then you need to put it inside quotation marks.

But when we read that data in using `read.csv()`, we haven't actually *stored* it to be able to work with.

To do that we need an extra step.


## Storing data in variables

A **variable** is an object that stores information in your code, in the same way as a spreadsheet cell might store different pieces of information in a spreadsheet.

Just as spreadsheet cells are given names like "A1" or "C3" (and [you can also give them custom names](https://support.microsoft.com/en-us/office/define-and-use-names-in-formulas-4d0f13ac-53b7-422e-afd2-abd7ff379c64)), variables are given names too.

Variables in coding, however, can contain much more complex information than can be stored in one spreadsheet cell. For example, a variable can contain a whole table, or a whole column of values, as well as just a single number or string of text.

The first time that you create a variable you need to give it a name, and **assign** a value to the variable.

This is done by putting the name of your variable, followed by the characters `<-` (think of it as an arrow pointing left, at that name), followed by what you want to store in that variable. Sometimes the value you want to store is very obvious, like so:

`myname <- "Paul"`

In this example the string of letters "Paul" is assigned to the variable called `myname`.

But if you are using code to generate information - like when importing data - you can also assign the **results** of some code to a variable, like so:

`yourvariable <- read.csv('yourdata.csv')`

In that example the results of applying the `read.csv()` function to the file 'yourdata.csv' are assigned to the variable called `yourvariable`.

The bit that says `yourvariable` is just a name that you choose. It can be almost anything, but choose something meaningful.

To **store** the gender pay gap, then, we need to adapt the code to read like this:

```{r}
genderpaygapdata <- read.csv("UK Gender Pay Gap Data - 2019 to 2020.csv")
```

In the code above we've added `genderpaygapdata <- ` to assign the results of that code to a new variable called `genderpaygapdata`.

You can tell that data has been stored in a variable in RStudio by looking in the *Environment* area in the upper right corner. Your new variable should now be listed there - and any new variables will be added there too. Data in particular will be ordered to appear at the top of the list.


## Variable types

When you import a CSV or other spreadsheet into R, it is stored in a particular *type* of variable, called a **data frame**. Data frames are basically tables - it is a type of variable that has rows and columns - but there is one important distinction to note: 

In a data frame a column can only contain *one type* of information. For example: all numbers, or all text. In spreadsheets, of course, you might have a column where some cells contain numbers and others contain text but that's not possible in a data frame: instead, a column with such a mix will be treated as if *all* the cells contained text. We will talk more about this distinction later.

First, now that we have the data stored in R, we can start to explore the different types of data within it.

A quick way to do that is with another function: `str()`.

The `str()` function tells you the **structure** of your data frame - including types of information. The one ingredient it needs is the name of the data frame, like so:

```{r}
str(genderpaygapdata)
```

Running this function will give you ("return" to you) a line telling you what type of object it is ("data frame") and how many rows and columns there are ("5775 obs. of  25 variables"). Note that rows are called "observations" and columns are called "variables".

This is followed by a series of lines with details on each column. The first column, for example, is `$ EmployerName: Factor w/ 5772 levels`, and the fifth column is `$ DiffMeanHourlyPercent: num`.

What does this mean?

* The dollar sign indicates that we are dealing with a column, or 'variable' - more on this later.
* Then we have the name of that column: EmployerName, for example
* We then have `Factor` or `num`. The `num` means it's a **numeric** column: `DiffMeanHourlyPercent`, for example, contains only numbers. But `Factor` needs some more explanation...

### Explaining factors

Factors are basically columns of categorical data. A column with only "male" or "female", for example, is a classic example of a factor. As [Hadley Wickham puts it in the book R for Data Science](https://r4ds.had.co.nz/factors.html):

> "In R, factors are used to work with categorical variables, variables that have a fixed and known set of possible values. They are also useful when you want to display character vectors in a non-alphabetical order.

> "Historically, factors were much easier to work with than characters. As a result, many of the functions in base R automatically convert characters to factors. This means that factors often crop up in places where they’re not actually helpful."




